---
title: "Abundance Modelling"
author: "Kwasi Wrensford"
date: "2025-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
library(vroom)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(forcats)
library(lubridate)
library(MASS)
library(hms)
library(unmarked)
library(camtrapR)
library(overlap)
library(stars)
library(sf)
library(terra)
library(rstan)
library(Rcpp)
library(stringr)
library(mgcv)
library(cowplot)
library(gridExtra)
library(ubms)
library(spAbundance)
library(spOccupancy)
library(MCMCvis)
library(forestplot)
library(purrr)
library(sjPlot)
library(sjmisc)
library(ggeffects)
library(cmdstanr)

```

```{r}
getwd()
```

#Formatting count data

```{r}

coyotecounts <- as.matrix(read.csv("data/coyotecounts.csv"))
lynxcounts <- as.matrix(read.csv("data/lynxcounts.csv"))
martencounts <- as.matrix(read.csv("data/martencounts.csv"))
fishercounts <- as.matrix(read.csv("data/fishercounts.csv"))
redfoxcounts <- as.matrix(read.csv("data/redfoxcounts.csv"))
```

#Load in covariate data

```{r}
covars <- read.csv("data/covars.csv")
```

```{r}
cor.test(scale(covars$linear), scale(covars$abmi_linear))

covars$PC1<-scale(covars$PC1)
covars$PC2<-scale(covars$PC2)
covars$hf<-scale(covars$hf)
covars$baited <- as.numeric(covars$baited)
```

#Formatting count data into one row per site
```{r}

site_covars <- covars %>%
  group_by(site) %>%
  slice(1) %>%   # keep the first camera per site
  ungroup()


make_site_level_mats <- function(Y, covars, occ_prefix = "o") {
  library(dplyr)
  library(tidyr)
  library(stringr)
  
  ### ---------------------------------------------------------
  ### 0. Fix duplicated occasion names BEFORE tibble conversion
  ### ---------------------------------------------------------
  
  # Extract original occasion columns
  occ_cols <- grep(paste0("^", occ_prefix), colnames(Y), value = TRUE)
  
  # Count how many times each base name appears
  occ_base <- sub(paste0("^", occ_prefix), "", occ_cols)
  K <- length(unique(occ_base))          # occasions per year
  total_K <- length(occ_cols)            # total columns
  n_years <- total_K / K                 # infer number of years
  
  if (total_K %% K != 0) {
    stop("Occasion columns do not divide evenly into years. Check your matrix.")
  }
  
  # Build new unique names: y1_o1, y1_o2, ..., y2_o1, y2_o2, ...
  year_index <- rep(seq_len(n_years), each = K)
  occ_index  <- rep(seq_len(K), times = n_years)
  
  new_names <- paste0("y", year_index, "_o", occ_index)
  
  # Apply new names BEFORE tibble conversion
  colnames(Y)[colnames(Y) %in% occ_cols] <- new_names
  
  ### ---------------------------------------------------------
  ### 1. Now safely convert to tibble
  ### ---------------------------------------------------------
  Y <- as_tibble(Y)
  covars <- as_tibble(covars)
  
  ### ---------------------------------------------------------
  ### 2. Standardize covariate names
  ### ---------------------------------------------------------
  covars_std <- covars %>%
    rename(
      camera = location,
      site   = site,
      baited = baited
    )
  
  ### ---------------------------------------------------------
  ### 3. Build long-format table
  ### ---------------------------------------------------------
  dat_long <- Y %>%
    mutate(
      camera = covars_std$camera,
      site   = covars_std$site
    ) %>%
    pivot_longer(
      cols = all_of(new_names),
      names_to = "occ_full",
      values_to = "count"
    ) %>%
    filter(!is.na(count)) %>%
    mutate(
      year = as.integer(str_extract(occ_full, "(?<=y)\\d+(?=_)")),
      occ  = as.integer(str_extract(occ_full, "(?<=_o)\\d+"))
    ) %>%
    arrange(site, camera, year, occ) %>%
    group_by(site) %>%
    mutate(rep_id = row_number()) %>%
    ungroup()
  
  ### ---------------------------------------------------------
  ### 4. Add baited status
  ### ---------------------------------------------------------
  bait_df <- covars_std %>% dplyr::select(site, camera, baited)
  
  dat_long <- dat_long %>%
    left_join(bait_df, by = c("site", "camera"))
  
  ### ---------------------------------------------------------
  ### 5. Build wide matrices
  ### ---------------------------------------------------------
  
  # Count matrix
  y_wide <- dat_long %>%
    dplyr::select(site, rep_id, count) %>%
    pivot_wider(names_from = rep_id, values_from = count) %>%
    arrange(site)
  
  y_mat <- as.matrix(y_wide[,-1])
  rownames(y_mat) <- y_wide$site
  
  # Bait matrix
  bait_wide <- dat_long %>%
    dplyr::select(site, rep_id, baited) %>%
    pivot_wider(names_from = rep_id, values_from = baited) %>%
    arrange(site)
  
  bait_mat <- as.matrix(bait_wide[,-1])
  rownames(bait_mat) <- bait_wide$site
  
  # Camera ID matrix
  cam_wide <- dat_long %>%
    dplyr::select(site, rep_id, camera) %>%
    pivot_wider(names_from = rep_id, values_from = camera) %>%
    arrange(site)
  
  cam_mat <- as.matrix(cam_wide[,-1])
  rownames(cam_mat) <- cam_wide$site
  
  ### ---------------------------------------------------------
  ### Return everything
  ### ---------------------------------------------------------
  list(
    y_mat = y_mat,
    bait_mat = bait_mat,
    cam_mat = cam_mat,
    dat_long = dat_long,
    K = K,
    n_years = n_years
  )
}
```

```{r}
##Coyote data
coyote_out <- make_site_level_mats(
  Y = coyotecounts,
  covars = covars
)

coyote_y     <- coyote_out$y_mat
coyote_bait  <- coyote_out$bait_mat
coyote_camid <- coyote_out$cam_mat

##Red Fox
redfox_out <- make_site_level_mats(
  Y = redfoxcounts,
  covars = covars
)

redfox_y     <- redfox_out$y_mat
redfox_bait  <- redfox_out$bait_mat
redfox_camid <- redfox_out$cam_mat

##Lynx
lynx_out <- make_site_level_mats(
  Y = lynxcounts,
  covars = covars
)

lynx_y     <- lynx_out$y_mat
lynx_bait  <- lynx_out$bait_mat
lynx_camid <- lynx_out$cam_mat

##Martenlynx
marten_out <- make_site_level_mats(
  Y = martencounts,
  covars = covars
)

marten_y     <- marten_out$y_mat
marten_bait  <- marten_out$bait_mat
marten_camid <- marten_out$cam_mat

##Fisher
fisher_out <- make_site_level_mats(
  Y = fishercounts,
  covars = covars
)

fisher_y     <- fisher_out$y_mat
fisher_bait  <- fisher_out$bait_mat
fisher_camid <- fisher_out$cam_mat
```

#Visualizing overlap
```{r}
count_list <- list(
  coyote = coyotecounts,
  lynx   = lynxcounts,
  marten = martencounts,
  fisher = fishercounts,
  redfox = redfoxcounts
)

site_id <- covars$location

sites_species <- c(
  coyote = sum(rowSums(coyotecounts, na.rm = TRUE) > 0),
  lynx   = sum(rowSums(lynxcounts, na.rm = TRUE) > 0),
  marten = sum(rowSums(martencounts, na.rm = TRUE) > 0),
  fisher = sum(rowSums(fishercounts, na.rm = TRUE) > 0),
  redfox = sum(rowSums(fishercounts, na.rm = TRUE) > 0)
)

site_counts <- map_df(names(count_list), function(sp) {
  mat <- count_list[[sp]]
  data.frame(
    site = site_id,
    species = sp,
    detections = rowSums(mat, na.rm = TRUE)
  )
})

ggplot(site_counts, aes(x = site, y = detections)) +
  geom_col() +
  facet_wrap(~ species, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Detections per Site for Each Species",
    x = "Site ID",
    y = "Total Detections"
  )
```

```{r}
presence_list <- lapply(count_list, function(mat) {
  as.numeric(rowSums(mat, na.rm = TRUE) > 0)
})

presence_df <- data.frame(
  site = site_id,
  coyote = presence_list$coyote,
  lynx   = presence_list$lynx,
  marten   = presence_list$marten,
  fisher = presence_list$fisher,
  redfox = presence_list$redfox
)

# get species names
sp <- names(count_list)

# function to count sites where sp occurs alone
alone_counts <- sapply(sp, function(s) {
  this_present  <- presence_df[[s]] == 1
  others_absent <- rowSums(presence_df[sp[sp != s]]) == 0
  sum(this_present & others_absent)
})

alone_counts

pairwise_counts <- function(matA, matB, spA, spB) {

  A_present <- rowSums(matA, na.rm =TRUE) > 0
  B_present <- rowSums(matB, na.rm =TRUE) > 0

  data.frame(
    sp1 = spA,
    sp2 = spB,
    both_present = sum(A_present & B_present),
    only_sp1     = sum(A_present & !B_present),
    only_sp2     = sum(!A_present & B_present)
  )
}

out1 <- pairwise_counts(coyotecounts, lynxcounts,   "coyote", "lynx")
out2 <- pairwise_counts(coyotecounts, martencounts, "coyote", "marten")
out3 <- pairwise_counts(coyotecounts, fishercounts, "coyote", "fisher")
out4 <- pairwise_counts(coyotecounts, redfoxcounts, "coyote", "redfox")
out5 <- pairwise_counts(lynxcounts,   martencounts, "lynx",   "marten")
out6 <- pairwise_counts(lynxcounts,   fishercounts, "lynx",   "fisher")
out7 <- pairwise_counts(lynxcounts,   fishercounts, "lynx",   "redfox")
out8 <- pairwise_counts(martencounts, fishercounts, "marten", "fisher")
out9 <- pairwise_counts(martencounts, fishercounts, "marten", "redfox")
out10 <- pairwise_counts(martencounts, fishercounts, "fisher", "redfox")

pairwise_summary <- rbind(out1, out2, out3, out4, out5, out6, out7, out8, out9, out10)
pairwise_summary

```

```{r}
species <- names(count_list)

pairs <- t(combn(species, 2))
colnames(pairs) <- c("sp1", "sp2")

overlap_results <- apply(pairs, 1, function(sp) {
  sum(presence_df[[sp[1]]] == 1 & presence_df[[sp[2]]] == 1)
})

overlap_df <- data.frame(
  sp1 = pairs[,1],
  sp2 = pairs[,2],
  overlap_sites = overlap_results
)

ggplot(overlap_df, aes(sp1, sp2, fill = overlap_sites)) +
  geom_tile() +
  geom_text(aes(label = overlap_sites)) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Site Overlap Between Species")

```



##Frequentist N-mixture in unmarked
```{r}
umf_coyote <- unmarkedFramePCount(
  y = coyotecounts,
  siteCovs = covars
)

coyote_nmix <- pcount(
formula = ~ bait + (1|camera) ~ 
  PC1 * hf + PC2 * hf,
data = umf_coyote,
mixture = "P"
)

coyote_nmix
plotEffects(coyote_nmix, "state", "PC1")
plotEffects(coyote_nmix, "state", "PC2")
plotEffects(coyote_nmix, "state", "hf")
plotEffects(coyote_nmix, "det", "baited")

coyote_lambda <- predict(coyote_nmix, type = "state")

covars$coyote_abund <- coyote_lambda$Predicted

lm_coyote <- lm(coyote_abund~hf, data = covars)
summary(lm_coyote)

covars$coyote_resid <- resid(lm_coyote)
cor.test(covars$coyote_resid, covars$hf)
```

#Coyote disaggregated human footprint
```{r}
#Linear features
coyote_nmix_linear <- pcount(
formula = ~ baited ~ 
  PC1 * linear + PC2 * linear,
data = umf_coyote,
mixture = "P"
)

coyote_nmix_linear
plotEffects(coyote_nmix_linear, "state", "PC1")
plotEffects(coyote_nmix_linear, "state", "PC2")
plotEffects(coyote_nmix_linear, "state", "linear")
plotEffects(coyote_nmix_linear, "det", "baited")
```
```{r}
#ABMI Linear features
coyote_nmix_linearabmi <- pcount(
formula = ~ baited ~ 
  PC1 * scale(abmi_linear) + PC2 * scale(abmi_linear),
data = umf_coyote,
mixture = "P"
)

coyote_nmix_linearabmi
plotEffects(coyote_nmix_linearabmi, "state", "PC1")
plotEffects(coyote_nmix_linearabmi, "state", "PC2")
plotEffects(coyote_nmix_linearabmi, "state", "abmi_linear")
plotEffects(coyote_nmix_linearabmi, "det", "baited")
```

```{r}
#Industry
coyote_nmix_industrial <- pcount(
formula = ~ baited ~ 
  PC1 * industrial + PC2 * industrial,
data = umf_coyote,
mixture = "P"
)

coyote_nmix_industrial
plotEffects(coyote_nmix_industrial, "state", "PC1")
plotEffects(coyote_nmix_industrial, "state", "PC2")
plotEffects(coyote_nmix_industrial, "state", "industrial")
plotEffects(coyote_nmix_industrial, "det", "baited")
```
```{r}
#Industry
coyote_nmix_agriculture <- pcount(
formula = ~ baited ~ 
  PC1 * agriculture + PC2 * agriculture,
data = umf_coyote,
mixture = "P"
)

coyote_nmix_agriculture
plotEffects(coyote_nmix_agriculture, "state", "PC1")
plotEffects(coyote_nmix_agriculture, "state", "PC2")
plotEffects(coyote_nmix_agriculture, "state", "agriculture")
plotEffects(coyote_nmix_agriculture, "det", "baited")
```

#Lynx nmix abundance in unmarked
```{r}
umf_lynx <- unmarkedFramePCount(
  y = lynxcounts,
  siteCovs = covars
)

lynx_nmix <- pcount(
formula = ~ baited ~ 
  PC1 * hf + PC2 * hf + scale(coyote_resid),
data = umf_lynx,
mixture = "P"
)

lynx_nmix
plotEffects(lynx_nmix, type = "state", covariate = "PC1")
plotEffects(lynx_nmix, type = "state", covariate = "PC2")
plotEffects(lynx_nmix, type = "state", covariate = "hf")
plotEffects(lynx_nmix, type = "state", covariate = "coyote_resid")
plotEffects(lynx_nmix, "det", "baited")

```
#Marten nmix abundance in unmarked
```{r}
umf_marten <- unmarkedFramePCount(
  y = martencounts,
  siteCovs = covars
)

marten_nmix <- pcount(
formula = ~ baited ~ 
  PC1 * hf + PC2 * hf + scale(coyote_resid),
data = umf_marten,
mixture = "P"
)

marten_nmix
plotEffects(marten_nmix, type = "state", covariate = "PC1")
plotEffects(marten_nmix, type = "state", covariate = "PC2")
plotEffects(marten_nmix, type = "state", covariate = "hf")
plotEffects(marten_nmix, type = "state", covariate = "coyote_resid")
plotEffects(marten_nmix, "det", "baited")
```

#Fisher nmix abundance in unmarked
```{r}
umf_fisher <- unmarkedFramePCount(
  y = fishercounts,
  siteCovs = covars
)

fisher_nmix <- pcount(
formula = ~ baited ~ 
  PC1 * hf + PC2 * hf + coyote_resid,
data = umf_fisher,
mixture = "P"
)

fisher_nmix
plotEffects(fisher_nmix, type = "state", covariate = "PC1")
plotEffects(fisher_nmix, type = "state", covariate = "PC2")
plotEffects(fisher_nmix, type = "state", covariate = "hf")
plotEffects(fisher_nmix, type = "state", covariate = "coyote_resid")
plotEffects(fisher_nmix, "det", "baited")
```

#Red Fox nmix abundance in unmarked
```{r}
umf_redfox <- unmarkedFramePCount(
  y = redfoxcounts,
  siteCovs = covars
)

redfox_nmix <- pcount(
formula = ~ baited ~ 
  PC1 * hf + PC2 * hf + coyote_resid,
data = umf_redfox,
mixture = "P"
)

redfox_nmix
plotEffects(redfox_nmix, type = "state", covariate = "PC1")
plotEffects(redfox_nmix, type = "state", covariate = "PC2")
plotEffects(redfox_nmix, type = "state", covariate = "hf")
plotEffects(redfox_nmix, type = "state", covariate = "coyote_resid")
plotEffects(redfox_nmix, "det", "baited")

```

#Stan N-mixture using ubms
```{r}
##Coyote
umf_coyote <- unmarkedFramePCount(
  y = coyote_y,
  siteCovs = site_covars,
  obsCovs = list(bait = coyote_bait, camera = coyote_camid)
)

coyote_nmix_ubms <- stan_pcount(formula =  ~ bait 
                                ~ PC1 * hf + PC2 * hf,
                     data = umf_coyote,
                     chains = 3,
                     iter = 500,
                     cores = 3)

coyote_nmix_ubms
plot_effects(coyote_nmix_ubms, submodel = "state")
plot_effects(coyote_nmix_ubms, submodel = "det")


traceplot(coyote_nmix_ubms, pars=c("beta_state", "beta_det"))

plot_posteriors(coyote_nmix_ubms)

coyote_gof<-gof(coyote_nmix_ubms, draws=100)
coyote_gof
plot(coyote_gof)
```

```{r}
###Predict interaction plot function
plot_interaction_ubms <- function(model,
                                  focal,
                                  moderator,
                                  species = NULL,
                                  submodel = "state",
                                  data,
                                  mod_probs = c(0.25, 0.75),
                                  n = 1000,
                                  ylab = NULL,
                                  xlab = NULL,
                                  title = NULL,
                                  col_low = "darkgreen",
                                  col_high = "darkorange3") {
  
  # 1. Extract moderator quantiles
  mod_vals <- quantile(data[[moderator]], probs = mod_probs, na.rm = TRUE)
  
  # 2. Build prediction grids
  make_grid <- function(mod_value) {
    grid <- data.frame(
      focal = seq(min(data[[focal]], na.rm = TRUE),
                  max(data[[focal]], na.rm = TRUE),
                  length.out = n)
    )
    names(grid)[1] <- focal
    
    # Add moderator
    grid[[moderator]] <- mod_value
    
    # Fix all other covariates at their means
    other_covs <- setdiff(names(data), c(focal, moderator))
    for (v in other_covs) {
      grid[[v]] <- mean(data[[v]], na.rm = TRUE)
    }
    grid
  }
  
  grid_low <- make_grid(mod_vals[1])
  grid_high <- make_grid(mod_vals[2])
  
  # 3. Predict
  pred_low <- ubms::predict(model,
                            submodel = submodel,
                            species = species,
                            newdata = grid_low)
  
  pred_high <- ubms::predict(model,
                             submodel = submodel,
                             species = species,
                             newdata = grid_high)
  
  # 4. Combine into a plotting dataframe
  df <- data.frame(
    cov = grid_low[[focal]],
    pred_lo = pred_low$Predicted,
    lower_lo = pred_low$`2.5%`,
    upper_lo = pred_low$`97.5%`,
    pred_hi = pred_high$Predicted,
    lower_hi = pred_high$`2.5%`,
    upper_hi = pred_high$`97.5%`
  )
  
  # 5. Labels
  if (is.null(ylab)) ylab <- paste("Predicted", species)
  if (is.null(xlab)) xlab <- focal
  if (is.null(title)) title <- paste("Effect of", focal, "x", moderator)
  
  # 6. Plot
  library(ggplot2)
  ggplot(df, aes(x = cov)) +
    geom_ribbon(aes(ymin = lower_lo, ymax = upper_lo),
                fill = col_low, alpha = 0.2) +
    geom_line(aes(y = pred_lo), color = col_low, linewidth = 1) +
    geom_ribbon(aes(ymin = lower_hi, ymax = upper_hi),
                fill = col_high, alpha = 0.2) +
    geom_line(aes(y = pred_hi), color = col_high, linewidth = 1) +
    labs(y = ylab, x = xlab, title = title) +
    theme_classic()
}
```

```{r}
coyote_pc1_plot <- plot_interaction_ubms(
  model = coyote_nmix_ubms,
  focal = "PC1",
  moderator = "hf",
  species = "coyote",
  data = site_covars,
  ylab = "Abundance",
  xlab = "Climate PC1",
  title = "Coyote abundance vs PC1 at low/high HF"
)

coyote_pc1_plot
```

```{r}
coyote_pc2_plot <- plot_interaction_ubms(
  model = coyote_nmix_ubms,
  focal = "PC2",
  moderator = "hf",
  species = "coyote",
  data = site_covars,
  ylab = "Abundance",
  xlab = "Climate PC2",
  title = "Coyote abundance vs PC2 at low/high HF"
)

coyote_pc2_plot
```

##Red Fox
```{r}
##Red Fox
umf_redfox <- unmarkedFramePCount(
  y = redfox_y,
  siteCovs = site_covars,
  obsCovs = list(bait = redfox_bait, camera = redfox_camid)
)

redfox_nmix_stan <- stan_pcount(formula =  ~ bait
                                ~ PC1 * hf + PC2 * hf,
                     data = umf_redfox,
                     chains = 3,
                     iter = 500,
                     cores = 3)

redfox_nmix_stan
plot_effects(redfox_nmix_stan, submodel = "state")
plot_effects(redfox_nmix_stan, submodel = "det")

traceplot(redfox_nmix_stan, pars=c("beta_state", "beta_det"))
plot_residuals(redfox_nmix_stan, submodel = "state", covariate = "hf")

plot_posteriors(redfox_nmix_stan)

redfox_gof<-gof(redfox_nmix_stan, draws=100)
redfox_gof
plot(redfox_gof)
```

```{r}
redfox_pc1_plot <- plot_interaction_ubms(
  model = redfox_nmix_stan,
  focal = "PC1",
  moderator = "hf",
  species = "red fox",
  data = site_covars,
  ylab = "Abundance",
  xlab = "Climate PC1",
  title = "Red Fox abundance vs PC1 at low/high HF"
)

redfox_pc1_plot
```

```{r}
redfox_pc2_plot <- plot_interaction_ubms(
  model = redfox_nmix_stan,
  focal = "PC2",
  moderator = "hf",
  species = "red fox",
  data = site_covars,
  ylab = "Abundance",
  xlab = "Climate PC2",
  title = "Red Fox abundance vs PC2 at low/high HF"
)

redfox_pc2_plot
```

#Lynx
```{r}
##Lynx
umf_lynx <- unmarkedFramePCount(
  y = lynx_y,
  siteCovs = site_covars,
  obsCovs = list(bait = lynx_bait, camera = lynx_camid)
)

lynx_nmix_stan <- stan_pcount(formula =  ~ bait
                                ~ PC1 * hf + PC2 * hf,
                     data = umf_lynx,
                     chains = 3,
                     iter = 500,
                     cores = 3)

lynx_nmix_stan
plot_effects(lynx_nmix_stan, submodel = "state")
plot_effects(lynx_nmix_stan, submodel = "det")

traceplot(lynx_nmix_stan, pars=c("beta_state", "beta_det"))

plot_posteriors(lynx_nmix_stan)

lynx_gof<-gof(lynx_nmix_stan, draws=100)
lynx_gof
plot(lynx_gof)
```

```{r}
lynx_pc1_plot <- plot_interaction_ubms(
  model = lynx_nmix_stan,
  focal = "PC1",
  moderator = "hf",
  species = "lynx",
  data = site_covars,
  ylab = "Abundance",
  xlab = "Climate PC1",
  title = "Lynx abundance vs PC1 at low/high HF"
)

lynx_pc1_plot
```

```{r}
lynx_pc2_plot <- plot_interaction_ubms(
  model = lynx_nmix_stan,
  focal = "PC2",
  moderator = "hf",
  species = "lynx",
  data = site_covars,
  ylab = "Abundance",
  xlab = "Climate PC2",
  title = "Lynx abundance vs PC2 at low/high HF"
)

lynx_pc2_plot
```

##Marten
```{r}
##marten
umf_marten <- unmarkedFramePCount(
  y = marten_y,
  siteCovs = site_covars,
  obsCovs = list(bait = marten_bait, camera = marten_camid)
)

marten_nmix_stan <- stan_pcount(formula =  ~ bait
                                ~ PC1 * hf + PC2 * hf,
                     data = umf_marten,
                     chains = 3,
                     iter = 500,
                     cores = 3)

marten_nmix_stan
plot_effects(marten_nmix_stan, submodel = "state")
plot_effects(marten_nmix_stan, submodel = "det")

traceplot(marten_nmix_stan, pars=c("beta_state", "beta_det"))
plot_residuals(marten_nmix_stan, submodel = "state", covariate = "hf")

plot_posteriors(marten_nmix_stan)

marten_gof<-gof(marten_nmix_stan, draws=100)
marten_gof
plot(marten_gof)
```

```{r}
marten_pc1_plot <- plot_interaction_ubms(
  model = marten_nmix_stan,
  focal = "PC1",
  moderator = "hf",
  species = "marten",
  data = site_covars,
  ylab = "Abundance",
  xlab = "Climate PC1",
  title = "marten abundance vs PC1 at low/high HF"
)

marten_pc1_plot
```

```{r}
marten_pc2_plot <- plot_interaction_ubms(
  model = marten_nmix_stan,
  focal = "PC2",
  moderator = "hf",
  species = "marten",
  data = site_covars,
  ylab = "Abundance",
  xlab = "Climate PC2",
  title = "marten abundance vs PC2 at low/high HF"
)

marten_pc2_plot
```

##Fisher NMix Models
```{r}
##Fisher
umf_fisher <- unmarkedFramePCount(
  y = fisher_y,
  siteCovs = site_covars,
  obsCovs = list(bait = fisher_bait, camera = fisher_camid)
)

fisher_nmix_stan <- stan_pcount(formula =  ~ bait
                                ~ PC1 * hf + PC2 * hf,
                     data = umf_fisher,
                     chains = 3,
                     iter = 500,
                     cores = 3)

fisher_nmix_stan
plot_effects(fisher_nmix_stan, submodel = "state")
plot_effects(fisher_nmix_stan, submodel = "det")

traceplot(fisher_nmix_stan, pars=c("beta_state", "beta_det"))
plot_residuals(fisher_nmix_stan, submodel = "state", covariate = "hf")

plot_posteriors(fisher_nmix_stan)

fisher_gof<-gof(fisher_nmix_stan, draws=100)
fisher_gof
plot(fisher_gof)
```

```{r}
fisher_pc1_plot <- plot_interaction_ubms(
  model = fisher_nmix_stan,
  focal = "PC1",
  moderator = "hf",
  species = "fisher",
  data = site_covars,
  ylab = "Abundance",
  xlab = "Climate PC1",
  title = "fisher abundance vs PC1 at low/high HF"
)

fisher_pc1_plot
```

```{r}
fisher_pc2_plot <- plot_interaction_ubms(
  model = fisher_nmix_stan,
  focal = "PC2",
  moderator = "hf",
  species = "fisher",
  data = site_covars,
  ylab = "Abundance",
  xlab = "Climate PC2",
  title = "fisher abundance vs PC2 at low/high HF"
)

fisher_pc2_plot
```

#Custom N-mixture in Stan
```{r}
#Format data
abund <- as.matrix(coyotecounts)   # cameras × occasions
K      <- ncol(abund)              # number of occasions

site_id   <- covars$site           # site per camera
sites     <- sort(unique(site_id))
I         <- length(sites)         # number of sites

# cameras per site
J_tab <- table(site_id)
J     <- as.integer(J_tab[match(sites, names(J_tab))])
max_J <- max(J)
max_K <- K

y      <- array(0L, dim = c(I, max_J, max_K))
use_y  <- array(0L, dim = c(I, max_J, max_K))
baited_arr <- matrix(0L, nrow = I, ncol = max_J)

clim_pc1 <- numeric(I)
clim_pc2 <- numeric(I)
human_fp <- numeric(I)

for (i in seq_along(sites)) {
  this_site <- sites[i]
  cams_i    <- which(site_id == this_site)   # row indices in covars/coyotecounts
  J_i       <- length(cams_i)

  # site-level covariates: same for all cameras at this site
  clim_pc1[i] <- covars$PC1[cams_i[1]]
  clim_pc2[i] <- covars$PC2[cams_i[1]]
  human_fp[i] <- covars$hf[cams_i[1]]

  for (j in seq_len(J_i)) {
    row_idx <- cams_i[j]

    # camera-level detection covariate
    baited_arr[i, j] <- as.integer(covars$baited[row_idx])

    for (k in seq_len(max_K)) {
      val <- abund[row_idx, k]
      if (!is.na(val)) {
        y[i, j, k]     <- as.integer(val)
        use_y[i, j, k] <- 1L
      }
    }
  }
}


stan_data <- list(
  I       = I,
  J       = J,
  max_J   = max_J,
  max_K   = max_K,
  y       = y,
  use_y   = use_y,
  clim_pc1 = clim_pc1,
  clim_pc2 = clim_pc2,
  human_fp = human_fp,
  baited   = baited_arr,
  K        = 100  # or some upper bound for N
)

```

```{r}
mod <- cmdstan_model(
  "C:/Users/Gaynor Lab/Documents/abmi_coyotes/coyote_nmix.stan"
)

str(stan_data, max.level = 1)

stan_data$J

i <- 1
j <- 1
cbind(
  y      = stan_data$y[i, j, ],
  use_y  = stan_data$use_y[i, j, ]
)

head(stan_data$clim_pc1)
head(stan_data$clim_pc2)
head(stan_data$human_fp)

stan_data$baited[1, 1:stan_data$J[2]]

stan_data$K
```
```{r}
system.time(
  fit_test <- mod$sample(
    data = stan_data,
    chains = 1,
    parallel_chains = 1,
    iter_warmup = 5,
    iter_sampling = 5,
    refresh = 1
  )
)

```

```{r}
fit_test <- mod$sample(
  data = stan_data,
  chains = 1,
  parallel_chains = 1,
  iter_warmup = 200,
  iter_sampling = 200,
  adapt_delta = 0.99,
  max_treedepth = 12,
  refresh = 20
)

fit_test$diagnostic_summary()[, c("variable", "rhat", "ess_bulk", "ess_tail", "divergent__", "treedepth__")]


```

```{r}
fit <- mod$sample(
  data = stan_data,
  chains = 2,
  parallel_chains = 2,
  iter_warmup = 300,
  iter_sampling = 300,
  adapt_delta = 0.9,
  max_treedepth = 12,
  refresh = 50
)

```
