---
title: "Multi-Season Occupancy"
author: "Kwasi Wrensford"
date: "2024-12-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(vroom)
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(lubridate)
library(hms)
library(unmarked)
library(camtrapR)
library(overlap)
library(spOccupancy)
library(stars)
library(sf)
library(terra)
set.seed(1996)
```



#Building a multiseason model for Alberta coyotes

##Loading in and cleaning data

Start by loading in camera data (should already have data downloaded locally and stored in gitignore)

```{r}
# Camera Reports
cameras<-unique(study_main_reports$location)

all_main_reports <- vroom::vroom("data/ABMI EH 2014-2019 Main Reports.csv")
all_image_reports <- vroom::vroom("data/ABMI EH 2014-2019 Image Reports.csv")

head(all_main_reports)
head(all_image_reports)

##Create column for year of report
all_main_reports$year <- lubridate::year(all_main_reports$image_date_time)
all_image_reports$year <- lubridate::year(all_image_reports$image_date_time)

##Filter reports to 2015-2019 range
study_image_reports<-all_image_reports %>%
  filter(year > 2014 & year < 2020)

study_main_reports<-all_main_reports %>%
  filter(year > 2014 & year < 2020)

sort(unique(all_main_reports$species_common_name))

##Use image reports file to extract camera operation dates for each location
camera_operation_dates_v2<-study_image_reports %>%
  group_by(year, project_id, location) %>%
  summarize(
    start_date_time = min(image_date_time),
    end_date_time = max(image_date_time)
  )

camera_operation_dates_v2$start_date_time<-lubridate::as_date(camera_operation_dates_v2$start_date_time)
camera_operation_dates_v2$end_date_time<-lubridate::as_date(camera_operation_dates_v2$end_date_time)
```


##Simulated data approach to troubleshoot building model
Simulate a dataset for troubleshooting code.
```{r}
n_sites <- 10
n_seasons <- 5
n_surveys <- 25 #In the real model this will be days

#Simulateed site locations
site_data <- data.frame(
  site = 1:n_sites,
  latitude = runif(n_sites, 35, 45),
  longitude = (runif(n_sites, -110, -100))
)
```

```{r}
site_sf <- st_as_sf(site_data, coords = c("longitude", "latitude"), crs = 4326)
site_sf_utm <- st_transform(site_sf, crs = 32613)  # UTM Zone 13N
site_data$utm_easting <- st_coordinates(site_sf_utm)[, 1]
site_data$utm_northing <- st_coordinates(site_sf_utm)[, 2]
```

```{r}
site_data$human_footprint <- runif(n_sites, 0, 100)  # Human footprint index (0-100 scale)
site_data$winter_intensity <- rnorm(n_sites, mean = -10, sd = 5)  # Avg winter temp (°C)
```

Simulate occupancy
```{r}
occ_prob <- plogis(0.5 + 0.1 * site_data$human_footprint - 0.2 * site_data$winter_intensity)

z <- matrix(rbinom(n_sites * n_seasons, 1, occ_prob), nrow = n_sites, ncol = n_seasons)
```

Simulate detection
```{r}
effort <- array(runif(n_sites * n_surveys * n_seasons, 1, 10), dim = c(n_sites, n_surveys, n_seasons))

det_prob <- plogis(-0.5 + 0.2 * effort)

# Simulate "Day of Survey" (random survey day within a season)
set.seed(123) 
day <- array(sample(1:30, n_sites * n_surveys * n_seasons, replace = TRUE), 
             dim = c(n_sites, n_surveys, n_seasons))
# Simulate "Time of Day" (e.g., morning = 0, evening = 1)
tod <- array(sample(0:1, n_sites * n_surveys * n_seasons, replace = TRUE), 
             dim = c(n_sites, n_surveys, n_seasons))

det.covs = list(
    day = day,
    tod = tod
   )

y <- array(0, dim = c(n_sites, n_surveys, n_seasons))# Initialize with 0

for (i in 1:n_sites) {
  for (j in 1:n_seasons) {
    for (k in 1:n_surveys) {
      y[i, k, j] <- rbinom(1, 1, det_prob[i, k, j] * z[i, j])  # Detection only if species is present
    }
  }
}
```

```{r}
occ_data <-list(
  y = y,
  occ.covs = site_data[, c("latitude", "human_footprint", "winter_intensity")],
  det.covs = det.covs,
  coords = as.matrix(site_data[,c("utm_easting", "utm_northing")])
)
```

Simple model to start
```{r}
model <- tPGOcc(
  occ.formula = ~ latitude + human_footprint + winter_intensity,  # Include environmental covariates
  det.formula = ~ day + tod,  # Detection model: depends on effort
  data = occ_data,
  n.batch = 1000,
  batch.length = 25,
  n.burn = 500,
  n.thin = 1,
  n.chains = 3
)

# Model summary
summary(model)
```

Define covariate formulas
```{r}
occ.formula <- ~year + latitude + human_footprint + winter_intensity
det.formula <- ~day + tod
```

Set initial values
```{r}
```

Set priors
```{r}
```

Fit model
```{r}
model <- tPGOcc(
  occ.formula = ~ latitude + human_footprint + winter_intensity,  # Include environmental covariates
  det.formula = ~ day + tod,  # Detection model: depends on effort
  data = occ_data,
  n.batch = 1000,
  batch.length = 25,
  n.burn = 500,
  n.thin = 1,
  n.chains = 3
)

# Model summary
summary(model)
```


##Real Data
Format data for spOccupancy, build 3 dimensional array of detections with site, primary time period (years), secondary time period (days)

```{r}
coyotes <- carnivores %>%
  filter(species_common_name == "Coyote")
coyote
coyotes$image_date <- date(coyotes$image_date_time)
coyotes$image_time <- as_hms(coyotes$image_date_time)
coyotes.long <- coyotes %>%
  group_by(location, year, image_date) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  glimpse()
```

```{r}
detection_history <- coyotes %>%
  group_by(location, year, image_date) %>%
  summarise(detection = 1, .groups = "drop") %>%
  complete(location, year, image_date, fill = list(detection = 0))
```

```{r}
location_list <- study_main_reports %>%
  distinct(location) %>%
  arrange(location)

year_list <- unique(coyotes$year)

detection_array <- detection_history %>%
  pivot_wider(names_from = image_date, values_from = detection, values_fill = 0) %>%
  select(-year) %>%
  as.matrix ()

num_locations <- length(location_list)
num_days <-ncol(detection_array)
num_years<-length(year_list)
detection_3d <-array(
  as.numeric(detection_array[, -1]),
  dim = c(num_locations, num_days, num_years),
  dimnames = list(location = location_list, day = 1:num_days, year = year_list)
)
```

##Covariates

###Coordinates

extract coordinates for all cam locations (will need to do this with unbuffered coordinates)

```{r}
cam.coordinates<-study_main_reports%>%
  select(location, latitude, longitude)%>%
  distinct(location, .keep_all = TRUE)

colnames(cam.coordinates) <- c("Location", "Latitude", "Longitude")
cam.coordinates<-as.data.frame(cam.coordinates)

cam.coordinates.11n <- cam.coordinates %>%
  filter(Longitude >= -120 & Longitude < -114)

cam.coordinates.12n <- cam.coordinates %>%
  filter(Longitude >= -114 & Longitude < -108)
```

convert to decimal coordinates to UTM for use in spOccupancy

```{r}
cam.coordinates.sf.11n <- st_as_sf(cam.coordinates.11n, coords = c("Longitude", "Latitude"), crs = 4326)
print(cam.coordinates.sf.11n)

cam.coordinates.sf.12n <- st_as_sf(cam.coordinates.12n, coords = c("Longitude", "Latitude"), crs = 4326)
print(cam.coordinates.sf.12n)
```

Convert to UTC (Split into two regions since Alberta straddles two UTC zones)

```{r}
#UTM zone 11n
cam.coordinates.11n.utm <- st_transform(cam.coordinates.sf.11n, crs = 32611)

#UTM zone 12n
cam.coordinates.12n.utm <- st_transform(cam.coordinates.sf.12n, crs = 32612)

```

Extract the UTM coordinates, and recombine to get a single dataframe containing location, lat, long, easting(x), and northing(y)

```{r}
utm.11n.coordinates <- st_coordinates(cam.coordinates.11n.utm)
utm.12n.coordinates <- st_coordinates(cam.coordinates.12n.utm)

# Recombine utm datasets
cam.coordinates.11n.combined<-cbind(cam.coordinates.11n, utm.11n.coordinates)
cam.coordinates.12n.combined<-cbind(cam.coordinates.12n, utm.12n.coordinates)

cam.coordinates.utm <- rbind(cam.coordinates.11n.combined, cam.coordinates.12n.combined)
utm.coordinates <- cam.coordinates.utm %>%
  arrange(Location) %>%
  select(X, Y)
```

###Latitude

```{r}
cam.latitude <- study_main_reports %>%
  select(location, latitude) %>%
  distinct(location, .keep_all = TRUE)

cam.latitude.occ<-cam.latitude %>%
  arrange(location) %>%
  select(latitude)
```

###Camera Effort (Detection covariate)



###Climate

```{r}

```

###Human Footprint

```{r}

```

###Habitat Characteristics

```{r}
```

##Package covariates and occupancy data together into list object

```{r}
occ_covs <- array(
  cam.latitude.occ
)
occ_data <- list(
  y = detection_3d,
  occ.covs = latitude,
  det.covs = 
  coords = utm.coordinates
)
```

##Fit occupancy model for coyote data

Dry run of simple model
```{r}
model.test <- stPGOcc(
  occ.formula = ~ Latitude,
  det.formula = ~ effort,
  data = occ_data,
  n.batch = 1000,
  bath.length = 25,
  n.buurn = 500,
  n.thin - 1,
  n.chains = 3
)
```
Define occupancy formula and detection formula

```{r}
coyote.sp.occ.formula <-  ~scale(year) + scale(lat)
coyote.sp.det.formula <- scale(day) + I(scale(day)^2 + scale(tod))
```

Specify initial values for all model parameters

```{r}
z.inits <- apply(coyote.data, c(1, 2), function(a) as.numeric(sum(a, na.rm = TRUE) > 0))

dist.hbef <- dist(coyote.data$coords)
coyote.sp.inits <- list(beta = 0, 
                        alpha = 0,  sigma.sq.psi = 1,
                        z = z.inits)
```

Specify priors

```{r}
coyote.sp.priors <- list(beta.normal = list(mean = 0, var = 2.72), 
                        alpha.normal = list(mean = 0, var = 2.72), 
                        sigma.sq.psi.ig = list(a = 0.1, b = 0.1))
```

Below, using the camtrapR package, we create a matrix of detection histories that can be used in an occupancy model object

```{r}
# camtrapR occupancy modeling workflow

##Camera operation matrix
camera_operation_dates_filtered<-camera_operation_dates_v2%>%
  filter(!is.na(start_date_time))

camera_operation_dates_filtered<-camera_operation_dates_filtered%>%
  mutate(start_date_time = as.character(start_date_time))%>%
  mutate(end_date_time = as.character(end_date_time))

##Start and end dates for deployments per year
camera_deployment_dates<-camera_operation_dates_filtered %>%
  group_by(location, year)%>%
  summarize(
    start_date_time = min(start_date_time),
    end_date_time = max (end_date_time),
    .groups = "drop"
  )

##Idenitfy gaps or periods of inactivity within a camera's deployment "problems"
problem_matrix <-camera_operation_dates_filtered%>%
  arrange(location, year, start_date_time)%>%
  group_by(location, year)%>%
  mutate(
    Problem1_from = lag(end_date_time) + 1,
    Problem1_to = start_date_time -1
  ) %>%
  filter(!is.na(Problem1_from)& Problem1_from <= Problem1_to)%>%
  select(location, year, Problem1_from, Problem1_to)

##Combine problems with deployment data
camera_deployment_problems<-camera_deployment_dates%>%
  left_join(problem_matrix, by = c("location", "year"))%>%
              arrange(location, year, start_date_time, Problem1_from)

##Convert dates to character to be compatible with camptrapR
camera_deployment_problems<-camera_deployment_problems%>%
  mutate(start_date_time = as.character(start_date_time))%>%
  mutate(end_date_time = as.character(end_date_time))

##Filter cameras where start and end time are identical
camera_deployment_problems<-camera_deployment_problems%>%
  filter(start_date_time != end_date_time)

#Build a camera operation matrix to fill with detections
camera_operation_matrix<-cameraOperation(camera_deployment_problems,
                                         stationCol = "location",
                                         sessionCol = "year",
                                         setupCol = "start_date_time",
                                         retrievalCol = "end_date_time",
                                         writecsv = FALSE,
                                         hasProblems = TRUE,
                                         dateFormat = "ymd_HMS"
                                         )
##Camera detection history
datetime_check<-as.POSIXct(study_main_reports$image_date_time, format = ymd_HMS)
invalid_entries<-study_main_reports[is.na(datetime_check),]
invalid_entries

###Daylight savings time!! 4 coyote observations fall in the DST hour, so shifting to 3:00 to compensate (I'll figure out a more elegant way to do this later)

##entries 255396,255397,1882366,7326233

study_main_reports[255396,9] <- as.POSIXct("2015-03-08 3:03:29", format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
study_main_reports[255397,9]<- as.POSIXct("2015-03-08 3:03:49", format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
study_main_reports[1882366,9]<- as.POSIXct("2015-03-08 3:00:01", format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
study_main_reports[7326233,9]<- as.POSIXct("2018-03-11 3:15:00", format = "%Y-%m-%d %H:%M:%S", tz = "UTC")

coyotedetectionhistory<-detectionHistory(recordTable = study_main_reports,
                                         camOp = camera_operation_matrix,
                                         stationCol = "location",
                                         speciesCol = "species_common_name",
                                         recordDateTimeCol = "image_date_time",
                                         recordDateTimeFormat = "ymd_HMS",
                                         species = "Coyote",
                                         occasionLength = 1,
                                         day1 = "station",
                                         datesAsOccasionNames = FALSE,
                                         includeEffort = TRUE,
                                         scaleEffort = FALSE,
                                         timeZone = "America/Edmonton",
                                         unmarkedMultFrameInput = TRUE
                                         )
coyotedetections<-coyotedetectionhistory$detection_history
```
