---
title: "formattingabmidata"
output: html_document
editor: visual 
editor_options: 
  chunk_output_type: console
---

```{r}
library(vroom)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggfortify)
library(ggrepel)
library(forcats)
library(lubridate)
library(unmarked)
library(camtrapR)
library(overlap)
library(ggmap)
library(reshape2)
library(corrplot)
library(terra)
library(mapview)
library(sf)
library(exactextractr)
library(xml2)
library(geodata)
library(viridis)
library(scales)
library(colorspace)
```

##Read in and format camera data

```{r}
# Camera Reports
all_main_reports <- vroom::vroom("data/ABMI EH 2014-2019 Main Reports.csv")
all_image_reports <- vroom::vroom("data/ABMI EH 2014-2019 Image Reports.csv")

camlocations <- all_main_reports %>%
  distinct(location, .keep_all = TRUE)

##Create column for year of report
all_main_reports$year <- lubridate::year(all_main_reports$image_date_time)
all_image_reports$year <- lubridate::year(all_image_reports$image_date_time)

##Create column for month
all_main_reports$month <- lubridate::month(all_main_reports$image_date_time)
all_image_reports$month <- lubridate::month(all_image_reports$image_date_time)

##Filter reports to 2015-2019 range
study_image_reports<-all_image_reports %>%
  filter(year > 2014 & year < 2020)

study_main_reports<-all_main_reports %>%
  filter(year > 2014 & year < 2020)

##Filter to Apr - July for closure

study_main_reports<-study_main_reports %>%
  filter(month > 3 & month < 8)

study_image_reports<-study_image_reports %>%
  filter(month > 3 & month < 8)

##Filter data for only mesos
study_main_reports <- study_main_reports %>%
  filter(species_common_name == "Coyote" 
         | species_common_name == "Canada Lynx"
         | species_common_name == "Red Fox"
         | species_common_name == "Marten"
         | species_common_name == "Fisher")

#Camera effort data
total_operation_days<- vroom::vroom("data/Total Days of Operations ABMI EH 2014-2020.csv")

study_operation_days <- total_operation_days %>%
  filter(!project %in% c(
    "ABMI Ecosystem Health 2014",
    "ABMI Ecosystem Health 2014",
    "ABMI Ecosystem Health 2020",
    "ABMI Ecosystem Health 2021",
    "ABMI Ecosystem Health 2022"
    ))

camera_operation_dates <- vroom::vroom("data/Operation Dates ABmi EH 2014-2018.csv")
```

##Covariate data

```{r}
#Load in human footprint dataset, buffered and unbuffered
ehs_humanfootprint<-vroom::vroom("data/Point Level Vegetation HF for Kwasi.csv")
ehs_humanfootprint_buffered<-vroom::vroom("data/150m Buffer Vegetation HF for Kwasi.csv")

#Combine buffered and unbuffered dataset
buffered_renamed <- ehs_humanfootprint_buffered %>%
  rename_with(~ paste0(.x, "_buf"),
              .cols = -location)

ehs_humanfootprint_merged <- ehs_humanfootprint %>%
  left_join(buffered_renamed, by = "location")
```

##Make sure covariate locations match detection locations

```{r}
#Extract cam locations from image reports
camlocations <- all_main_reports %>%
  distinct(location, .keep_all = TRUE)

# Join the two datasets by Station
check_coords <- ehs_humanfootprint_merged %>%
  left_join(camlocations %>% dplyr::select(location, location_id, latitude_cam = latitude, longitude_cam = longitude),
            by = "location")

# Flag mismatches
check_coords <- check_coords %>%
  mutate(
    lat_mismatch = Lat != latitude_cam,
    lon_mismatch = Long != longitude_cam
  )

# View rows with mismatches
filter(check_coords, lat_mismatch | lon_mismatch)

```

```{r}
##Replace incorrect coordinates in covariate dataset with correct ones from cam reports
covariates_fixed <- check_coords %>%
  mutate(
    Latitude  = ifelse(lat_mismatch, latitude_cam, Lat),
    Longitude = ifelse(lon_mismatch, longitude_cam, Long)
  ) %>%
  dplyr::select(-latitude_cam, -longitude_cam, -lat_mismatch, -lon_mismatch, -Lat, -Long)
```


##Independent detections
```{r}
#Generate delta time (intervals between detection)
study_intervals <- study_main_reports %>%
  arrange(species_common_name, location, image_date_time) %>%
  group_by(species_common_name, location) %>%
  mutate(delta_time = as.numeric(difftime(image_date_time, lag(image_date_time), units = "mins"))) %>%
  ungroup()
```

```{r}
#Set independence threshold to 5 minutes
threshold <- 5

independent_detections <- study_intervals %>%
  arrange(species_common_name, location, image_date_time) %>%
  group_by(species_common_name, location) %>%
  mutate(
    independent = if_else(
      is.na(delta_time) | delta_time > threshold,
      TRUE,
      FALSE
    )
  ) %>%
  ungroup()

independent_events <- independent_detections %>%
  filter(independent)

independent_events %>%
  group_by(species_common_name) %>%
  summarize(
    n = n()
  )

study_main_reports <- independent_events
```

##Latitudinal and Natural Region Partitioning


```{r}
#Plotting camera locations by natural region
ggplot(covariates_fixed, aes(x = Longitude, y = Latitude, color = NR)) + 
  theme_minimal()+
  geom_point()

ggplot(covariates_fixed, aes(x = NR, fill = NSR)) + 
  theme_minimal()+
  geom_bar()


```

```{r}
#Identify how many of each species are observed in each Natural Region
study_main_reports <- study_main_reports %>%
  left_join(covariates_fixed %>% dplyr::select(location, NR, NSR),
            by = "location")

species_list <- unique(study_main_reports$species_common_name)

detections_summary <- study_main_reports %>%
  group_by(NSR, NR, species_common_name) %>%
  summarise(count = n(), .groups = "drop")

 region_colors <- c(
  "Rocky Mountain"  = "#1b9e77",
  "Boreal"          = "#d95f02",
  "Parkland"        = "#7570b3",
  "Grassland"       = "#e7298a",
  "Canadian Shield" = "#66a61e",
  "Foothills"       = "#e6ab02",
  "Unknown"         = "grey50"   # for NA
) 

 detections_summary <- detections_summary %>%
  mutate(
    NR_clean = ifelse(is.na(NR), "Unknown", NR),
    NSR = factor(NSR, levels = unique(NSR[order(NR_clean, NSR)]))
  )
 
 detections_plot <- detections_summary %>%
  mutate(NR_clean = ifelse(is.na(NR), "Unknown", NR)) %>%
  group_by(NR_clean) %>%
  mutate(
    subregion_color = lighten(
      region_colors[NR_clean],
      amount = seq(0, 0.4, length.out = n_distinct(NSR))
    )[as.numeric(factor(NSR))]
  ) %>%
  ungroup()


ggplot(detections_plot,
       aes(x = NR, y = count, fill = NR, color = NR)) +
  geom_col() +
  facet_wrap(~ species_common_name, scales = "free_y") +
  scale_color_manual(values = region_colors)+
  scale_fill_manual(values = region_colors) +
  labs(
    x = "Subregion",
    y = "Detections",
    fill = "Natural Region",
    title = "Detections per subregion (shaded by natural region)"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
#Group Natural SubRegions for Analyses

covariates_fixed$region <- with(covariates_fixed, case_when(
  
  # Group 1: Canadian Shield + Boreal except Dry Mixedgrass
  NR == "Canadian Shield" ~ "Shield+Boreal",
  NR == "Boreal" & NSR != "Dry Mixedwood" ~ "Shield+Boreal",
  
  # Group 2: Grassland + Parkland + Dry Mixedgrass
  NR %in% c("Grassland", "Parkland") ~ "Grassland+Parkland",
  NSR == "Dry Mixedwood" ~ "Grassland+Parkland",
  
  # Group 3: Rocky Mountain + Foothills
  NR %in% c("Rocky Mountain", "Foothills") ~ "Rocky+Foothills",
  
  TRUE ~ "Other"   # safety catch; should not be needed
))


table(covariates_fixed$region)

table(covariates_fixed$NSR, covariates_fixed$region)

covariates_fixed %>%
  distinct(region, NR, NSR) %>%
  arrange(region, NR, NSR) %>%
  group_by(region, NR) %>%
  summarise(NSR = paste(NSR, collapse = ", "), .groups = "drop") %>%
  group_by(region) %>%
  summarise(
    summary = paste0(NR, ": ", NSR, collapse = "\n  "),
    .groups = "drop"
  ) %>%
  mutate(summary = paste0(region, "\n  ", summary)) %>%
  pull(summary) %>%
  cat(sep = "\n\n")

##Map of Regions
ggplot(covariates_fixed, aes(x = Longitude, y = Latitude, color = region)) + 
  theme_minimal()+
  geom_point()

covariates_fixed <- covariates_fixed %>%
  mutate(
    # Grouping 0: original regions
    region_group0 = region,

    # Grouping 1: Combine Rocky+Foothills + Shield+Boreal
    region_group1 = case_when(
      region %in% c("Rocky+Foothills", "Shield+Boreal") ~ "Rocky/Shield Combined",
      region == "Grassland+Parkland" ~ "Grassland+Parkland",
      TRUE ~ NA_character_
    ),

    # Grouping 2: Drop Rocky+Foothills entirely
    region_group2 = case_when(
      region == "Shield+Boreal" ~ "Shield+Boreal",
      region == "Grassland+Parkland" ~ "Grassland+Parkland",
      TRUE ~ NA_character_
    )
  )

```

```{r}
#Species per region

det_with_region <- study_main_reports %>%
  left_join(covariates_fixed %>% select(location, region),
            by = "location")

det_summary <- det_with_region %>%
  count(species_common_name, region)

ggplot(det_summary, aes(x = region, y = n, fill = species_common_name)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(
    x = "Region",
    y = "Number of Detections",
    fill = "Species",
    title = "Detections per Region for All Focal Species"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

##Create detection and count matrices for each species

```{r}
# camtrapR occupancy modeling workflow

#Filter study image reports to locations in filtered main reports
study_camlocations <- unique(study_main_reports$location)

study_image_reports <- study_image_reports %>%
  filter(location %in% study_camlocations)

##Join study region data to detections
cov_filtered <- covariates_fixed %>%
  semi_join(study_main_reports, by = "location") %>%
  distinct(location, .keep_all = TRUE)

study_main_reports <- study_main_reports %>%
  left_join(cov_filtered, by = "location")


##Use image reports file to extract camera operation dates for each location
camera_operation_dates_v2<-study_image_reports %>%
  group_by(year, project_id, location) %>%
  summarize(
    start_date_time = min(image_date_time),
    end_date_time = max(image_date_time)
  )


##Build a Camera operation matrix
camera_operation_dates_filtered<-camera_operation_dates_v2%>%
  filter(!is.na(start_date_time))

#camera_operation_dates_filtered<-camera_operation_dates_filtered%>%
#  mutate(start_date_time = as.character(start_date_time))%>%
 # mutate(end_date_time = as.character(end_date_time))

##Start and end dates for deployments per year
camera_deployment_dates<-camera_operation_dates_filtered %>%
  group_by(location, year)%>%
  summarize(
    start_date_time = min(start_date_time),
    end_date_time = max (end_date_time),
    .groups = "drop"
  )

##Idenitfy gaps or periods of inactivity within a camera's deployment "problems"
problem_matrix <-camera_operation_dates_filtered%>%
  arrange(location, year, start_date_time)%>%
  group_by(location, year)%>%
  mutate(
    Problem1_from = lag(end_date_time) + 1,
    Problem1_to = start_date_time -1
  ) %>%
  filter(!is.na(Problem1_from)& Problem1_from <= Problem1_to)%>%
  dplyr::select(location, year, Problem1_from, Problem1_to)

##Combine problems with deployment data
camera_deployment_problems<-camera_deployment_dates%>%
  left_join(problem_matrix, by = c("location", "year"))%>%
              arrange(location, year, start_date_time, Problem1_from)

##Convert dates to character to be compatible with camptrapR
#camera_deployment_problems<-camera_deployment_problems%>%
#  mutate(start_date_time = as.character(start_date_time))%>%
#  mutate(end_date_time = as.character(end_date_time))

##Filter cameras where start and end time are identical
camera_deployment_problems<-camera_deployment_problems%>%
  filter(start_date_time != end_date_time)

#Build a camera operation matrix to fill with detections
camera_operation_matrix<-cameraOperation(camera_deployment_problems,
                                         stationCol = "location",
                                         sessionCol = "year",
                                         setupCol = "start_date_time",
                                         retrievalCol = "end_date_time",
                                         writecsv = FALSE,
                                         hasProblems = FALSE,
                                         dateFormat = "ymd_HMS"
                                         )

camera_operation_matrix_clean <- camtrapR:::deparseCamOpRownames(camera_operation_matrix)

rownames(camera_operation_matrix_clean) <- camera_operation_matrix_clean$station
camera_operation_matrix_clean$session <- NULL

camera_operation_matrix_clean <- as.matrix(camera_operation_matrix_clean)

##Camera detection history
datetime_check<-as.POSIXct(study_main_reports$image_date_time, format = ymd_HMS)
invalid_entries<-study_main_reports[is.na(datetime_check),]
invalid_entries

###Daylight savings time!! 4 coyote observations fall in the DST hour, so shifting to 3:00 to compensate (I'll figure out a more elegant way to do this later)

##entries 255396,255397,1882366,7326233

#study_main_reports[255396,9] <- as.POSIXct("2015-03-08 3:03:29", format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
#study_main_reports[255397,9]<- as.POSIXct("2015-03-08 3:03:49", format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
#study_main_reports[1882366,9]<- as.POSIXct("2015-03-08 3:00:01", format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
#study_main_reports[7326233,9]<- as.POSIXct("2018-03-11 3:15:00", format = "%Y-%m-%d %H:%M:%S", tz = "UTC")

##Create detection tables for each focal species with a week  long (7 day) occasion length

#Function

make_matrices <- function(det_df,
                          camop_df,
                          species,
                          region = NULL,
                          occasion_length = 7,
                          day1 = "station",
                          tz = "America/Edmonton") {
  
  #------------------------------------------------------------
  # 1. Filter by species
  #------------------------------------------------------------
  df <- det_df %>% 
    filter(species_common_name == species)
  
  if (nrow(df) == 0) {
    stop(paste("No records found for species:", species))
  }
  
  #------------------------------------------------------------
  # 2. Optional region filter
  #------------------------------------------------------------
  if (!is.null(region)) {
    
    # Check region exists
    if (!region %in% unique(det_df$region)) {
      stop(paste("Region", region, "not found in detection dataset"))
    }
    
    df <- df %>% filter(region == !!region)
    
    if (nrow(df) == 0) {
      stop(paste("No records for species", species, "in region", region))
    }
    
    # Subset camOp to only stations present in filtered detections
    camop_sub <- camop_df[rownames(camop_df) %in% df$location, ]
    
  } else {
    camop_sub <- camop_df
  }
  
  #------------------------------------------------------------
  # 3. Detection history (binary)
  #------------------------------------------------------------
  dh <- detectionHistory(
    recordTable = df,
    camOp = camop_sub,
    stationCol = "location",
    speciesCol = "species_common_name",
    recordDateTimeCol = "image_date_time",
    recordDateTimeFormat = "ymd_HMS",
    species = species,
    occasionLength = occasion_length,
    day1 = day1,
    datesAsOccasionNames = FALSE,
    includeEffort = TRUE,
    scaleEffort = FALSE,
    timeZone = tz,
    output = "binary",
    unmarkedMultFrameInput = TRUE
  )
  
  #------------------------------------------------------------
  # 4. Count matrix
  #------------------------------------------------------------
  cm <- detectionHistory(
    recordTable = df,
    camOp = camop_sub,
    stationCol = "location",
    speciesCol = "species_common_name",
    recordDateTimeCol = "image_date_time",
    recordDateTimeFormat = "ymd_HMS",
    species = species,
    occasionLength = occasion_length,
    day1 = day1,
    datesAsOccasionNames = FALSE,
    includeEffort = TRUE,
    scaleEffort = FALSE,
    timeZone = tz,
    output = "count",
    unmarkedMultFrameInput = TRUE
  )
  
  #------------------------------------------------------------
  # 5. Return both matrices
  #------------------------------------------------------------
  return(list(
    detection_matrix = dh$detection_history,
    count_matrix = cm$detection_history
  ))
}


coyote <- make_matrices(
  det_df = study_main_reports,
  camop_df = camera_operation_matrix,
  species = "Coyote"
)

coyote_boreal <- make_matrices(
  det_df = study_main_reports,
  camop_df = camera_operation_matrix_clean,
  species = "Coyote",
  region = "Shield+Boreal"
)

coyote_park <- make_matrices(
  det_df = study_main_reports,
  camop_df = camera_operation_matrix,
  species = "Coyote"
)

coyote$detection_matrix
coyote$count_matrix

redfox <- make_matrices(
  det_df = study_main_reports,
  camop_df = camera_operation_matrix,
  species = "Red Fox"
)

lynx <- make_matrices(
  det_df = study_main_reports,
  camop_df = camera_operation_matrix,
  species = "Canada Lynx"
)

marten <- make_matrices(
  det_df = study_main_reports,
  camop_df = camera_operation_matrix,
  species = "Marten"
)

fisher <- make_matrices(
  det_df = study_main_reports,
  camop_df = camera_operation_matrix,
  species = "Fisher"
)
```

##Align covariates location names with detection matrix locations (same number and order)

```{r}
covariates_aligned <- covariates_fixed %>%
  filter(location %in% row.names(coyote$detection_matrix)) %>%
  arrange(match(location, row.names(coyote$detection_matrix)))

# ensure same order
all(row.names(coyote$detection_matrix) == covariates_aligned$location)

sites_det <- rownames(coyote$detection_matrix)
sites_cov <- covariates_aligned$location

setdiff(sites_det, sites_cov)  # sites in detections but not in covariates
setdiff(sites_cov, sites_det)  # sites in covariates but not in detections

common_sites <- intersect(sites_det, sites_cov)

# filter detections
coyote$detection_matrix <- coyote$detection_matrix[common_sites, ]
coyote$count_matrix <- coyote$count_matrix[common_sites, ]
lynx$detection_matrix <- lynx$detection_matrix[common_sites, ]
lynx$count_matrix <- lynx$count_matrix[common_sites, ]
marten$detection_matrix <- marten$detection_matrix[common_sites, ]
marten$count_matrix <- marten$count_matrix[common_sites, ]
fisher$detection_matrix <- fisher$detection_matrix[common_sites, ]
fisher$count_matrix <- fisher$count_matrix[common_sites, ]
redfox$detection_matrix <- redfox$detection_matrix[common_sites, ]
redfox$count_matrix <- redfox$count_matrix[common_sites, ]

coyote$detection_matrix <- coyote$detection_matrix[rownames(coyote$detection_matrix) != "1095-SE",]
coyote$count_matrix <- coyote$count_matrix[rownames(coyote$count_matrix) != "1095-SE",]
lynx$detection_matrix <- lynx$detection_matrix[rownames(lynx$detection_matrix) != "1095-SE",]
lynx$count_matrix <- lynx$count_matrix[rownames(lynx$count_matrix) != "1095-SE",]
marten$detection_matrix <- marten$detection_matrix[rownames(marten$detection_matrix) != "1095-SE",]
marten$count_matrix <- marten$count_matrix[rownames(marten$count_matrix) != "1095-SE",]
fisher$detection_matrix <- fisher$detection_matrix[rownames(fisher$detection_matrix) != "1095-SE",]
fisher$count_matrix <- fisher$count_matrix[rownames(fisher$count_matrix) != "1095-SE",]
redfox$detection_matrix <- redfox$detection_matrix[rownames(redfox$detection_matrix) != "1095-SE",]
redfox$count_matrix <- redfox$count_matrix[rownames(redfox$count_matrix) != "1095-SE",]

# filter covariates
covariates_aligned <- covariates_aligned %>%
  filter(location %in% common_sites) %>%
  arrange(match(location, common_sites))

# Which sites are duplicated?
dup_sites <- covariates_aligned$location[duplicated(covariates_aligned$location)]
dup_sites

# Count how many times each site occurs
table(covariates_aligned$location)

covariates_unique <- covariates_aligned %>%
  distinct(location, .keep_all = TRUE)

identical(rownames(coyote$detection_matrix), covariates_unique$location)
identical(rownames(lynx$detection_matrix), covariates_unique$location)
identical(rownames(marten$detection_matrix), covariates_unique$location)
identical(rownames(fisher$detection_matrix), covariates_unique$location)
identical(rownames(redfox$detection_matrix), covariates_unique$location)
```

##Dimension reduction of climate data

```{r}
## Annual Climate variables
#1) Annual variables:

#Directly calculated annual variables:

#MAT              mean annual temperature (°C),

#MWMT           mean warmest month temperature (°C),

#MCMT            mean coldest month temperature (°C),

#MAP               mean annual precipitation (mm),

#AHM  annual heat-moisture index (MAT+10)/(MAP/1000))

#FFP                Frost free period

#PET                Potential evapotranspiration

##Filter out climate covariates

covariates_climate <- covariates_unique %>%
  dplyr::select(location, Latitude, Longitude, AHM, PET, FFP, MAP, MAT, MWMT, MCMT, region)

##Keep location info
covar_locations <- covariates_climate %>%
  dplyr::select(location, Latitude, Longitude, region)

##Climate variables only
climate_var <- covariates_climate %>%
    dplyr::select(-location, -Latitude, -Longitude, -region)

climate_scaled <- scale(climate_var)

##Run PCA
pca_res <- prcomp(climate_scaled, center = TRUE, scale. = TRUE)

plot(pca_res, type = "lines")

summary(pca_res)

biplot(pca_res)


##Extract PC's
climate_pcs <- as.data.frame(pca_res$x[, 1:2])
colnames(climate_pcs) <- c("PC1", "PC2")

# Loadings (variable contributions to PCs)
round(pca_res$rotation[, 1:2], 3)
load_df <- as.data.frame(pca_res$rotation[, 1:2])
load_df$label <- rownames(load_df)

# Bind with site info
climate_sites <- cbind(covar_locations, climate_pcs)

#Plot PC's
mult <- min(
  (max(autoplot(pca_res)$data$PC1) - min(autoplot(pca_res)$data$PC1)) /
    (max(load_df$PC1) - min(load_df$PC1)),
  (max(autoplot(pca_res)$data$PC2) - min(autoplot(pca_res)$data$PC2)) /
    (max(load_df$PC2) - min(load_df$PC2))
)

load_df[, c("PC1", "PC2")] <- load_df[, c("PC1", "PC2")] * mult

pc_plot <- autoplot(pca_res, data = climate_sites, color = "Latitude",
         loadings = TRUE, loadings.color = "black", 
         loadings.label = FALSE)+
  geom_point(aes(color = Latitude), size = 3) +
  scale_color_viridis_c(option = "plasma", direction = -1)+
  labs(x = "PC1 - Overall Climate",
       y = "PC2 - Aridity")+
  theme_classic() +
  theme(axis.title = element_text(size = 15))

pc_plot + 
  geom_text_repel(
    data = load_df,
    aes(x = PC1, y = PC2, label = label),
    fontface = "bold",
    size = 4,
    color = "black",
    box.padding = 0.5,
    point.padding = 0.3,
    segment.color = "grey40",
    force = 3,
    max.overlaps = Inf
)
```

##ABMI Linear Features
```{r}
linear_abmi <- covariates_unique %>%
  dplyr::select(location, Latitude, Longitude, TransmissionLine_buf, Pipeline_buf, SeismicLineNarrow_buf, SeismicLineWide_buf)

linear_abmi <- linear_abmi %>%
  mutate(abmi_linear = rowSums(.[4:7]))
```


##Canada Human Footprint index

```{r}
library(terra)
library(geodata)

# Load rasters (human footprint TIFF) and crop to alberta provincial boundary
hf_raster <- rast("data/chfi/cum_threat2020.02.18.tif")
roads_raster <- rast("data/chfi/roads.tif")
rail_raster <- rast("data/chfi/rail.tif")
built_raster <- rast("data/chfi/built.tif")
oil_gas_raster <- rast("data/chfi/oil_gas.tif")
forestry_raster <- rast("data/chfi/forestry_harvest.tif")
mines_raster <- rast("data/chfi/mines.tif")
crop_raster <- rast("data/chfi/crop.tif")
pasture_raster <- rast("data/chfi/pasture.tif")

roads_raster <- resample(roads_raster, hf_raster)
rail_raster <- resample(rail_raster, hf_raster)
built_raster <- resample(built_raster, hf_raster)
oil_gas_raster <- resample(oil_gas_raster, hf_raster)
forestry_raster <- resample(forestry_raster, hf_raster)
mines_raster <- resample(mines_raster, hf_raster)
crop_raster <- resample(crop_raster, hf_raster)
pasture_raster <- resample(pasture_raster, hf_raster)

hf_stack <- c(hf_raster, roads_raster, rail_raster, built_raster, oil_gas_raster, forestry_raster, mines_raster, crop_raster, pasture_raster)

names(hf_stack) <- c("hf", "roads", "rail", "built", "oil_gas", "forestry", "mines", "crop", "pasture")

##Obtain Canada and Alberta boundary shapefiles

url <- "https://geospatial.alberta.ca/titan/rest/services/boundary/goa_administrative_area/MapServer/0"

canada <- gadm("Canada", level = 1, path = "data/")
alberta <- canada[canada$NAME_1 == "Alberta", ]
alberta_proj <- project(alberta, hf_stack)

hf_alberta <- crop(hf_stack, alberta_proj)
hf_alberta_masked <- mask(hf_alberta, alberta_proj)

# Convert sites to spatial object
covar_locations <- covar_locations %>%
  filter(!is.na(Longitude) & !is.na(Latitude))

sites_hf <- st_as_sf(covar_locations, coords = c("Longitude", "Latitude"), crs = 4326)
sites_hf <- st_transform(sites_hf, crs(hf_alberta_masked))
sites_vect <- vect(sites_hf)

plot(hf_alberta_masked$hf)
plot(sites_vect, add = TRUE, col = "darkorange", cex = 0.5)

buffers <- st_buffer(sites_hf, dist = 5000)
buffers_vect <- vect(buffers)

#Linear Features
hf_linear <- hf_alberta_masked$rail + hf_alberta_masked$roads
names(hf_linear) <- "linear"

plot(hf_linear)
plot(sites_vect, add = TRUE, col = "darkorange", cex = 0.5)

#Industrial
hf_industrial <- hf_alberta_masked$oil_gas + hf_alberta_masked$mines + hf_alberta_masked$forestry
names(hf_industrial) <- "industrial"

plot(hf_industrial)
plot(sites_vect, add = TRUE, col = "darkorange", cex = 0.5)

#Agriculture
hf_agriculture <- hf_alberta_masked$crop + hf_alberta_masked$pasture
names(hf_agriculture) <- "agriculture"

plot(hf_agriculture)
plot(sites_vect, add = TRUE, col = "darkorange", cex = 0.5)

hf_vals <- terra::extract(hf_alberta_masked, buffers_vect, fun = mean, na.rm = TRUE)
linear_vals <- terra::extract(hf_linear, buffers_vect, fun = mean, na.rm = TRUE)
industrial_vals <- terra::extract(hf_industrial, buffers_vect, fun = mean, na.rm = TRUE)
agriculture_vals <- terra::extract(hf_agriculture, buffers_vect, fun = mean, na.rm = TRUE)

covar_locations_hf <- bind_cols(covar_locations, hf_vals[,-1])

```

##Compiling Covariates into dataframe
```{r}
##Build covariate dataframe
climate_sites <- climate_sites %>%
  filter(location != ("1095-SE"))

covars <- climate_sites %>%
  left_join(covar_locations_hf)

covars <- covars %>%
  bind_cols(linear_vals[-1])%>%
  bind_cols(industrial_vals[-1])%>%
  bind_cols((agriculture_vals[-1]))

#Create columns for site id and camera orientation
covars <- covars %>% 
  separate(
    col = location,
    into = c("site", "cam_orientation"),
    sep = "-",     # split at the hyphen
    remove = FALSE # keep original column if helpful
  )

study_main_reports <- study_main_reports %>%
  separate(
    col = location,
    into = c("site", "cam_orientation"),
    sep = "-",     # split at the hyphen
    remove = FALSE # keep original column if helpful
  )

#Create column for baited and non baited
covars <- covars %>%
  mutate(
    baited = cam_orientation %in% c("NE", "SW")
  )
```

```{r}
#Add abmi linear feature data
covars <-covars %>%
  bind_cols(linear_abmi[8])

#Correlation
hf_corr_data <- covars %>%
  dplyr::select(hf, linear, industrial, agriculture)

corr_matrix <- cor(hf_corr_data, use = "complete.obs", method = "pearson")
print(corr_matrix)
corrplot::corrplot(corr_matrix, method = "color", type = "upper")

barplot(covars$linear)
barplot(covars$industrial)
barplot(covars$agriculture)

long_covars <- covars %>%
    dplyr::select(location, hf, linear, industrial, agriculture) %>%
    pivot_longer(cols = -location,
                 names_to = "covariate",
                 values_to = "value")

ggplot(long_covars, aes(x = value)) +
    geom_histogram(bins = 30) +
    facet_wrap(~ covariate, scales = "free") +
    theme_bw()
```

#PC1 variation across province
```{r}
sites_joined <- merge(sites_vect, coyote_covars, by = "location")

sites_joined$PC1<-as.numeric(sites_joined$PC1)
sites_joined$PC2<-as.numeric(sites_joined$PC2)
sites_joined$hf<-as.numeric(sites_joined$hf)
sites_joined$coyote_counts <-as.numeric(sites_joined$coyote_counts)

mapview(
  sites_joined,
  zcol = "PC1",         # variable used to color the points
  cex = "coyote_counts",                   # point size
  legend = TRUE, 
  col.regions = viridis(20), # continuous palette
  layer.name = "Climate PC1"
)

mapview(
  sites_joined,
  zcol = "PC2",         # variable used to color the points
  cex = "coyote_counts",                   # point size
  legend = TRUE, 
  col.regions = viridis(20), # continuous palette
  layer.name = "Climate PC2"
)

mapview(
  sites_joined,
  zcol = "hf",         # variable used to color the points
  cex = "coyote_counts",                   # point size
  legend = TRUE, 
  col.regions = viridis(20), # continuous palette
  layer.name = "Human footprint"
)
```
Covariate variation across regions
```{r}

# Basic theme
theme_map <- theme_minimal(base_size = 13) +
  theme(panel.grid = element_blank())

# PC1 map
ggplot(covars, aes(x = Longitude, y = Latitude, color = PC1)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  theme_map +
  labs(title = "Spatial Variation in Climate PC1",
       color = "PC1")

# PC2 map
ggplot(covars, aes(x = Longitude, y = Latitude, color = PC2)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_c(option = "viridis") +
  theme_map +
  labs(title = "Spatial Variation in Climate PC2",
       color = "PC2")

# Human footprint map
ggplot(covars, aes(x = Longitude, y = Latitude, color = hf)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_c(option = "magma") +
  theme_map +
  labs(title = "Spatial Variation in Human Footprint",
       color = "HF")

covars_long <- covars %>%
  select(Longitude, Latitude, PC1, PC2, hf) %>%
  pivot_longer(cols = c(PC1, PC2, hf),
               names_to = "variable",
               values_to = "value")

ggplot(covars_long, aes(x = Longitude, y = Latitude, color = value)) +
  geom_point(size = 1.8, alpha = 0.8) +
  scale_color_viridis_c() +
  facet_wrap(~ variable, ncol = 1, scales = "free") +
  theme_map +
  labs(title = "Spatial Variation in Climate PCs and Human Footprint",
       color = "Value")

covars_long2 <- covars %>%
  select(region, PC1, PC2, hf) %>%
  pivot_longer(cols = c(PC1, PC2, hf),
               names_to = "variable",
               values_to = "value")

ggplot(covars_long2, aes(x = region, y = value, fill = region)) +
  geom_violin(alpha = 0.6) +
  geom_boxplot(width = 0.2, outlier.size = 0.8) +
  facet_wrap(~ variable, scales = "free_y") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(title = "Region-Level Variation in Climate PCs and Human Footprint",
       x = "Region", y = "Value")

```

```{r}
#Comparing covariates across natural regions and analysis groups

plot_data1 <- covars %>%
  filter(!is.na(region_group1)) %>%
  select(region_group1, PC1, PC2, hf) %>%
  pivot_longer(cols = c(PC1, PC2, hf),
               names_to = "variable",
               values_to = "value")

plot_data2 <- covars %>%
  filter(!is.na(region_group2)) %>%
  select(region_group2, PC1, PC2, hf) %>%
  pivot_longer(cols = c(PC1, PC2, hf),
               names_to = "variable",
               values_to = "value")
p1 <- ggplot(plot_data1, aes(x = region_group1, y = value, fill = region_group1)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15, outlier.size = 0.8) +
  facet_wrap(~ variable, scales = "free_y") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        legend.position = "none") +
  labs(title = "A) Rocky+Foothills + Shield+Boreal Combined vs Grassland+Parkland",
       x = "Region Group", y = "Value")

p2 <- ggplot(plot_data2, aes(x = region_group2, y = value, fill = region_group2)) +
  geom_violin(alpha = 0.6, trim = FALSE) +
  geom_boxplot(width = 0.15, outlier.size = 0.8) +
  facet_wrap(~ variable, scales = "free_y") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        legend.position = "none") +
  labs(title = "B) Shield+Boreal vs Grassland+Parkland (Rocky+Foothills Removed)",
       x = "Region Group", y = "Value")

library(patchwork)

combined_plot <- p1 / p2  # stacked panels
combined_plot

site_counts_by_group <- covars %>%
  select(site, region_group0, region_group1, region_group2) %>%
  distinct() %>%
  pivot_longer(cols = starts_with("region_group"),
               names_to = "grouping",
               values_to = "region_group") %>%
  filter(!is.na(region_group)) %>%
  count(grouping, region_group, name = "n_sites")

dets_joined <- study_main_reports %>%
  left_join(covars %>% select(site, region_group0, region_group1, region_group2),
            by = "site")

detection_counts_by_group <- dets_joined %>%
  pivot_longer(cols = starts_with("region_group"),
               names_to = "grouping",
               values_to = "region_group") %>%
  filter(!is.na(region_group)) %>%
  count(grouping, region_group, name = "n_detections")

p_sites <- ggplot(site_counts_by_group,
                  aes(x = region_group, y = n_sites, fill = region_group)) +
  geom_col() +
  facet_wrap(~ grouping, scales = "free_x") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        legend.position = "none") +
  labs(title = "Number of Sites per Region Grouping",
       x = "Region Group", y = "Number of Sites")

p_dets <- ggplot(detection_counts_by_group,
                 aes(x = region_group, y = n_detections, fill = region_group)) +
  geom_col() +
  facet_wrap(~ grouping, scales = "free_x") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        legend.position = "none") +
  labs(title = "Number of Detections per Region Grouping",
       x = "Region Group", y = "Number of Detections")

combined_counts <- p_sites / p_dets
combined_counts


```

#Covariate data exploration
```{r}
#Correlation between human footprint and PC1
cor.test(covars$PC1, covars$hf)

summary(lm(data= covars, PC1 ~ hf))

#Correlation between human footprint and PC2
cor.test(covars$PC2, covars$hf)

#Correlation between latitude and PC1
cor.test(covars$Latitude, covars$PC1)

#Correlation between latitude and PC2
cor.test(covars$Latitude, covars$PC2)
```


#Write CSV files

```{r}
write.csv(covariates_unique, file = "data/ehscovars.csv", row.names = FALSE)
write.csv(covars, file = "data/covars.csv", row.names = FALSE)

write.csv(coyote$detection_matrix, file = "data/coyotedetections.csv", row.names = FALSE)
write.csv(lynx$detection_matrix, file = "data/lynxdetections.csv", row.names = FALSE)
write.csv(marten$detection_matrix, file = "data/martendetections.csv", row.names = FALSE)
write.csv(fisher$detection_matrix, file = "data/fisherdetections.csv", row.names = FALSE)
write.csv(redfox$detection_matrix, file = "data/redfoxdetections.csv", row.names = FALSE)

write.csv(coyote$count_matrix, file = "data/coyotecounts.csv", row.names = FALSE)
write.csv(lynx$count_matrix, file = "data/lynxcounts.csv", row.names = FALSE)
write.csv(marten$count_matrix, file = "data/martencounts.csv", row.names = FALSE)
write.csv(fisher$count_matrix, file = "data/fishercounts.csv", row.names = FALSE)
write.csv(redfox$count_matrix, file = "data/redfoxcounts.csv", row.names = FALSE)

```
