---
title: "Abundance Modelling"
author: "Kwasi Wrensford"
date: "2025-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
library(vroom)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(forcats)
library(lubridate)
library(MASS)
library(hms)
library(unmarked)
library(camtrapR)
library(overlap)
library(stars)
library(Rcpp)
library(stringr)
library(mgcv)
library(cowplot)
library(gridExtra)
library(ubms)
library(MCMCvis)
library(sjPlot)
library(sjmisc)
library(ggeffects)
library(cmdstanr)
library(rjags)
library(jagsUI)
```

#Formatting count data

```{r}

coyotecounts <- as.matrix(read.csv("data/coyotecounts.csv"))
coyotecounts_boreal <- as.matrix(read.csv("data/coyotecounts_boreal.csv"))
coyotecounts_park <- as.matrix(read.csv("data/coyotecounts_park.csv"))


redfoxcounts <- as.matrix(read.csv("data/redfoxcounts.csv"))
redfoxcounts_boreal <- as.matrix(read.csv("data/redfoxcounts_boreal.csv"))
redfoxcounts_park <- as.matrix(read.csv("data/redfoxcounts_park.csv"))

lynxcounts <- as.matrix(read.csv("data/lynxcounts.csv"))
lynxcounts_boreal <- as.matrix(read.csv("data/lynxcounts_boreal.csv"))

martencounts <- as.matrix(read.csv("data/martencounts.csv"))
martencounts_boreal <- as.matrix(read.csv("data/martencounts.csv"))

fishercounts <- as.matrix(read.csv("data/fishercounts.csv"))
fishercounts_boreal <- as.matrix(read.csv("data/fishercounts_boreal.csv"))



```

#Load in covariate data

```{r}
covars <- read.csv("data/covars.csv")
covars_boreal <- read.csv("data/covars_boreal.csv")
covars_park <- read.csv("data/covars_park.csv")
```

```{r}
#Scale data for all covariates
covars$PC1<-scale(covars$PC1)
covars$PC2<-scale(covars$PC2)
covars$hf<-scale(covars$hf)
covars$baited <- as.numeric(covars$baited)

covars_boreal$PC1<-scale(covars_boreal$PC1)
covars_boreal$PC2<-scale(covars_boreal$PC2)
covars_boreal$hf<-scale(covars_boreal$hf)
covars_boreal$baited <- as.numeric(covars_boreal$baited)

covars_park$PC1<-scale(covars_park$PC1)
covars_park$PC2<-scale(covars_park$PC2)
covars_park$hf<-scale(covars_park$hf)
covars_park$baited <- as.numeric(covars_park$baited)
```

```{r}
ggplot(covars, aes(PC1, fill = region))+
  geom_histogram()

ggplot(covars, aes(PC2, fill = region))+
  geom_histogram()

ggplot(covars, aes(hf, fill = region))+
  geom_histogram()
```

#Custom N-mixture in Stan
```{r}
#Simulate data for model troubleshooting
set.seed(123)

# core design
n_site    <- 150          # number of sites
min_cam   <- 2
max_cam   <- 4
T         <- 10           # sampling occasions per camera

# number of cameras per site (2–4)
J_site <- sample(min_cam:max_cam, n_site, replace = TRUE)

sum(J_site)  # total cameras

# site-level covariates
clim_pc1 <- rnorm(n_site, 0, 1)
clim_pc2 <- rnorm(n_site, 0, 1)
human_fp       <- rnorm(n_site, 0, 1)  # or runif if you prefer 0–1 scaling

# build camera-level index
site_id_cam <- rep(1:n_site, times = J_site)
n_cam       <- length(site_id_cam)

# baited vs non-baited at camera level (0/1)
bait_cam <- rbinom(n_cam, size = 1, prob = 0.5)

# abundance (lambda) model: log(lambda_i) = beta0 + beta1*pc1 + beta2*pc2 + beta3*hf
beta0 <- 0.5
beta1 <- 0.6
beta2 <- -0.3
beta3 <- 0.4

beta_int1 <- 0.5   # NEW: interaction PC1 × HF
beta_int2 <- -0.4  # NEW: interaction PC2 × HF

phi_true <- 2.0
psi_zero <- 0.15

# detection model: logit(p_ijk) = alpha0 + alpha1*bait_cam_j
alpha0 <- -1.2
alpha1 <- 1.0

sigma_cam <- 0.7
u_cam <- rnorm(n_cam, 0, sigma_cam)

eta_lambda <- beta0 +
  beta1 * clim_pc1 +
  beta2 * clim_pc2 +
  beta3 * human_fp +
  beta_int1 * clim_pc1 * human_fp +   # NEW
  beta_int2 * clim_pc2 * human_fp     # NEW

lambda_site <- exp(eta_lambda)

is_zero <- rbinom(n_site, 1, psi_zero)

N_nb <- rnbinom(n_site, size = phi_true, mu = lambda_site)

N_site <- ifelse(is_zero == 1, 0, N_nb)
summary(N_site)

N_site <- rpois(n_site, lambda_site)
summary(N_site)

inv_logit <- function(x) 1 / (1 + exp(-x))

eta_p_cam <- alpha0 + alpha1 * bait_cam + u_cam
p_cam     <- inv_logit(eta_p_cam)

max_J <- max(J_site)
y <- array(NA_integer_, dim = c(n_site, max_J, T))

cam_counter <- 1
for (i in 1:n_site) {
  Ji <- J_site[i]
  Ni <- N_site[i]
  
  for (j in 1:Ji) {
    p_ij <- p_cam[cam_counter]
    y[i, j, ] <- rbinom(T, size = Ni, prob = p_ij)
    cam_counter <- cam_counter + 1
  }
}

bait_mat <- matrix(0, n_site, max_J)
cam_counter <- 1
for (i in 1:n_site) {
  Ji <- J_site[i]
  bait_mat[i, 1:Ji] <- bait_cam[cam_counter:(cam_counter + Ji - 1)]
  cam_counter <- cam_counter + Ji
}

use_y <- ifelse(!is.na(y), 1L, 0L)

y_fixed <- y
y_fixed[is.na(y_fixed)] <- 0L


stan_data <- list(
  I        = n_site,
  J        = J_site,
  max_J    = max_J,
  max_K    = T,
  y        = y_fixed,
  use_y    = use_y,
  clim_pc1 = clim_pc1,
  clim_pc2 = clim_pc2,
  human_fp = human_fp,
  baited   = bait_mat,
  K        = max(N_site) + 20   # upper bound for N
)


```

```{r}

mod <- cmdstan_model(
  "coyote_nmix.stan"
)

fit_test <- mod$sample(
  data = stan_data,
  chains = 2,
  parallel_chains = 2,
  iter_warmup = 200,
  iter_sampling = 200,
  adapt_delta = 0.99,
  max_treedepth = 12,
  refresh = 20
)

fit_test$diagnostic_summary()[, c("variable", "rhat", "ess_bulk", "ess_tail", "divergent__", "treedepth__")]


```
```

```{r}
#Format data
abund <- as.matrix(coyotecounts)   # cameras × occasions
K      <- ncol(abund)              # number of occasions

site_id   <- covars$site           # site per camera
sites     <- sort(unique(site_id))
I         <- length(sites)         # number of sites

# cameras per site
J_tab <- table(site_id)
J     <- as.integer(J_tab[match(sites, names(J_tab))])
max_J <- max(J)
max_K <- K

y      <- array(0L, dim = c(I, max_J, max_K))
use_y  <- array(0L, dim = c(I, max_J, max_K))
baited_arr <- matrix(0L, nrow = I, ncol = max_J)

clim_pc1 <- numeric(I)
clim_pc2 <- numeric(I)
human_fp <- numeric(I)

for (i in seq_along(sites)) {
  this_site <- sites[i]
  cams_i    <- which(site_id == this_site)   # row indices in covars/coyotecounts
  J_i       <- length(cams_i)

  # site-level covariates: same for all cameras at this site
  clim_pc1[i] <- covars$PC1[cams_i[1]]
  clim_pc2[i] <- covars$PC2[cams_i[1]]
  human_fp[i] <- covars$hf[cams_i[1]]

  for (j in seq_len(J_i)) {
    row_idx <- cams_i[j]

    # camera-level detection covariate
    baited_arr[i, j] <- as.integer(covars$baited[row_idx])

    for (k in seq_len(max_K)) {
      val <- abund[row_idx, k]
      if (!is.na(val)) {
        y[i, j, k]     <- as.integer(val)
        use_y[i, j, k] <- 1L
      }
    }
  }
}


stan_data <- list(
  I       = I,
  J       = J,
  max_J   = max_J,
  max_K   = max_K,
  y       = y,
  use_y   = use_y,
  clim_pc1 = clim_pc1,
  clim_pc2 = clim_pc2,
  human_fp = human_fp,
  baited   = baited_arr,
  K        = 100  # or some upper bound for N
)

```

```{r}
mod <- cmdstan_model(
  "C:/Users/Gaynor Lab/Documents/abmi_coyotes/coyote_nmix.stan"
)

str(stan_data, max.level = 1)

stan_data$J

i <- 1
j <- 1
cbind(
  y      = stan_data$y[i, j, ],
  use_y  = stan_data$use_y[i, j, ]
)

head(stan_data$clim_pc1)
head(stan_data$clim_pc2)
head(stan_data$human_fp)

stan_data$baited[1, 1:stan_data$J[2]]

stan_data$K
```
```{r}
system.time(
  fit_test <- mod$sample(
    data = stan_data,
    chains = 1,
    parallel_chains = 1,
    iter_warmup = 5,
    iter_sampling = 5,
    refresh = 1
  )
)

```

```{r}
fit_test <- mod$sample(
  data = stan_data,
  chains = 1,
  parallel_chains = 1,
  iter_warmup = 200,
  iter_sampling = 200,
  adapt_delta = 0.99,
  max_treedepth = 12,
  refresh = 20
)

fit_test$diagnostic_summary()[, c("variable", "rhat", "ess_bulk", "ess_tail", "divergent__", "treedepth__")]


```

```{r}
fit <- mod$sample(
  data = stan_data,
  chains = 2,
  parallel_chains = 2,
  iter_warmup = 300,
  iter_sampling = 300,
  adapt_delta = 0.9,
  max_treedepth = 12,
  refresh = 50
)

```
